Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-09-29 ms/main.tex
sec:cor -> corollary; more small revisions

\begin{corollary}[Symmetric horizontal transmission] 
\label{sec:cor_equal_transmission}
If $T=T_A=T_B$, then cooperation will take over the population if
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \alpha \frac{T}{1-T} \;.
\end{equation}
\end{corollary}
Note that the right-hand side equals $\gamma_1$ when $T=T_A=T_B$.
This condition is obtained by setting $T=T_A=T_B$ in inequality \ref{eq:vert_hori_global_condition} and can be interpreted as a version of Hamilton's rule  ($c<b\cdot r$, inequality~\ref{eq:hamilton_rule}) or as a version of inequality~\ref{eq:fletcher2009}, where $\alpha T/(1-T)$ can be regarded as the \emph{effective relatedness} or \emph{effective assortment}, respectively.
\autoref{fig:results}a illustrates this condition.

@@

\begin{corollary}[Intermediate association of transmission and cooperation]\label{corollary:Intermediate_association}
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}

@@

If $T=T_A=T_B$ then $\beta_1=0$ and $\beta_3=-\beta_2=\alpha b vT -cv(1-T)$, 
and $f(\hat{p})$ becomes a quadratic polynomial:
\begin{equation} \label{eq:equal_horizontal_transmission}
  f(\hat{p}) = \hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].
\end{equation}
Clearly the only two equilibria are the fixations  $\hat{p} =  0$ and $\hat{p} = 1$.
These equilibria are locally stable if $f'(\hat{p})<0$ near the equilibrium (see Appendix~\autoref{sec:appendixB}), where
\begin{equation}
f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
\end{equation}
with
\begin{equation} \label{eq:derivative_of_phattag-phat}
\begin{aligned}
	f'(0) &=	\alpha bvT - cv(1-T), \\
	f'(1) &=	-\alpha bvT + cv(1-T).
\end{aligned}
\end{equation}
Therefore with symmetric horizontal transmission ($T_A=T_B$), fixation of the cooperative phenotype ($\hat{p}=1$) occurs under the same condition as Corollary \ref{sec:cor_equal_transmission}, namely inequality~\ref{eq:equal_transmission}.

@@

In general, this condition cannot be formulated in the form of Hamilton's rule due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, i.e., with $T=T_A=T_B$, these conditions reduce to a form of Hamilton's rule.
\\
\begin{corollary}[Necessary condition for fixation of cooperation]
\label{sec:cor_vhat}
  Fixation of cooperation is possible only if the vertical transmission rate is high enough,
  \begin{equation} \label{eq:fixation_of_cooperation_vertical_transmission_condition}
    \begin{aligned}
      v>\hat{v} = \frac{T_B - T_A}{1-T_A} \;.
    \end{aligned}
    \end{equation} 
\end{corollary}
%Note that $\sigma \frac{\partial \gamma_1}{\partial v} = \sigma (T_B-T_A)$ and $\sigma \frac{\partial \hat\gamma_2}{\partial v} = \sigma (T_A-T_B)(b \alpha T_B - (1+b)(1-T_A))$
Note that this condition does not guarantee fixation of cooperation. Rather, if this condition does not apply then cooperation cannot fix.
When horizontal transmission is biased for cooperation, $T_A>T_B$, cooperation can fix with any vertical transmission rate, i.e. $\hat{v}<0$.
In contrast, if the horizontal transmission is biased for defection, $T_A<T_B$,  cooperation can fix only if the vertical transmission rate is high enough: in this case oblique transmission can prevent fixation of cooperation, see \autoref{fig:result3}bd.
\\

In corollary~\autoref{corollary:Intermediate_association}, we examined the effect of social association on evolution of cooperation in the case of perfect vertical transmission ($v=1$).
A more general case can be stated for $0 < v \leq 1$.
Consider the social association boundaries 
\begin{equation}\begin{aligned}\label{eq:boundries_assortative_meeting_general_case}
  \hat{a}_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  \hat{a}_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A} \;.
\end{aligned}\end{equation}
Then the following applies.
\\

\begin{corollary}[Intermediate association of transmission and cooperation]\label{corollary:Intermediate_association}
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; \hat a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; \hat a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}
\autoref{fig:result3}cd illustrates this conditions.

@@

We find that increasing social association ($\alpha$) increases the opportunity for evolution of cooperation (corollary X, \autoref{fig:result2}c and \autoref{fig:result3}cd). 
Similarly, a horizontal transmission bias for cooperation ($T_A-T_B$) will also facilitate its evolution (corollary X, \autoref{fig:result2}}.
The situation with vertical/oblique transmission ($v$) is more complex. 
When there is horizontal transmission bias for cooperation ($T_A>T_B$), increasing the rate of oblique transmission ($1-v$) will facilitate the evolution of cooperation (\autoref{fig:result3}ac). However, when the bias if for defection ($T_A<T_B$), high rates of vertical transmission ($v$) are advantageous for cooperation. 
Nevertheless, to reach fixation of cooperation in the population, there must be a sufficient rate of vertical transmission ($v>\hat v$, Corollary~\autoref{sec:cor_vhat}).
% TODO carry on summarizing corollaries here

Our analytic conditions derived for a deterministic well-mixed model provides a good approximation to results of simulations of a complex spatial stochastic model  (\autoref{fig:spatial}).
In this spatial model, individuals can only interact with and transmit to their neighbors on a grid.
This demonstrates that social association between cooperation and transmission can arise in a structured population in which interactions are local.

>>>>

\begin{corollary}[Symmetric horizontal transmission] 
\label{corollary:equal_transmission}
If $T=T_A=T_B$, then cooperation will take over the population if
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \alpha \frac{T}{1-T} \;.
\end{equation}
\end{corollary}
Note that the right-hand side equals $\gamma_1$ when $T=T_A=T_B$.
This condition is obtained by setting $T=T_A=T_B$ in inequality \ref{eq:vert_hori_global_condition} and can be interpreted as a version of Hamilton's rule  ($c<b\cdot r$, inequality~\ref{eq:hamilton_rule}) or as a version of inequality~\ref{eq:fletcher2009}, where $\alpha T/(1-T)$ can be regarded as the \emph{effective relatedness} or \emph{effective assortment}, respectively.
\autoref{fig:results}a illustrates this condition.

@@

\begin{corollary}[Intermediate association of transmission and cooperation]\label{corollary:intermediate_association}
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}

@@

If $T=T_A=T_B$ then $\beta_1=0$ and $\beta_3=-\beta_2=\alpha b vT -cv(1-T)$, 
and $f(\hat{p})$ becomes a quadratic polynomial:
\begin{equation} \label{eq:equal_horizontal_transmission}
  f(\hat{p}) = \hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].
\end{equation}
Clearly the only two equilibria are the fixations  $\hat{p} =  0$ and $\hat{p} = 1$.
These equilibria are locally stable if $f'(\hat{p})<0$ near the equilibrium (see Appendix~\autoref{sec:appendixB}), where
\begin{equation}
f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
\end{equation}
with
\begin{equation} \label{eq:derivative_of_phattag-phat}
\begin{aligned}
	f'(0) &=	\alpha bvT - cv(1-T), \\
	f'(1) &=	-\alpha bvT + cv(1-T).
\end{aligned}
\end{equation}
Therefore with symmetric horizontal transmission ($T_A=T_B$), fixation of the cooperative phenotype ($\hat{p}=1$) occurs under the same condition as Corollary \ref{corollary:equal_transmission}, namely inequality~\ref{eq:equal_transmission}.

@@

In general, this condition cannot be formulated in the form of Hamilton's rule due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, i.e., with $T=T_A=T_B$, these conditions reduce to a form of Hamilton's rule.
\\
\begin{corollary}[Necessary condition for fixation of cooperation]
\label{corollary:vhat}
  Fixation of cooperation is possible only if the vertical transmission rate is high enough,
  \begin{equation} \label{eq:fixation_of_cooperation_vertical_transmission_condition}
    \begin{aligned}
      v>\hat{v} = \frac{T_B - T_A}{1-T_A} \;.
    \end{aligned}
    \end{equation} 
\end{corollary}
%Note that $\sigma \frac{\partial \gamma_1}{\partial v} = \sigma (T_B-T_A)$ and $\sigma \frac{\partial \hat\gamma_2}{\partial v} = \sigma (T_A-T_B)(b \alpha T_B - (1+b)(1-T_A))$
Note that this condition does not guarantee fixation of cooperation. Rather, if this condition does not apply then cooperation cannot fix.
When horizontal transmission is biased for cooperation, $T_A>T_B$, cooperation can fix with any vertical transmission rate, i.e. $\hat{v}<0$.
In contrast, if the horizontal transmission is biased for defection, $T_A<T_B$,  cooperation can fix only if the vertical transmission rate is high enough: in this case oblique transmission can prevent fixation of cooperation, see \autoref{fig:result3}bd.
\\

In Corollary~\autoref{corollary:intermediate_association}, we examined the effect of social association on evolution of cooperation in the case of perfect vertical transmission ($v=1$).
A more general case can be stated for $0 < v \leq 1$.
Consider the social association boundaries 
\begin{equation}\begin{aligned}\label{eq:boundries_assortative_meeting_general_case}
  \hat{a}_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  \hat{a}_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A} \;.
\end{aligned}\end{equation}
Then the following applies.
\\

\begin{corollary}[Intermediate association of transmission and cooperation]
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; \hat a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; \hat a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}
\autoref{fig:result3}cd illustrates this conditions.

@@

We find that increasing social association ($\alpha$) increases the opportunity for evolution of cooperation (corollary X, \autoref{fig:result2}c and \autoref{fig:result3}cd). 
Similarly, a horizontal transmission bias for cooperation ($T_A-T_B$) will also facilitate its evolution (corollary X, \autoref{fig:result2}).
The effect of vertical/oblique transmission is more complex. 
When there is horizontal transmission bias for cooperation ($T_A>T_B$), increasing the rate of oblique transmission ($1-v$) will facilitate the evolution of cooperation (\autoref{fig:result3}ac). However, when the bias is for defection ($T_A<T_B$), high rates of vertical transmission ($v$) are advantageous for cooperation. 
Nevertheless, to reach fixation of cooperation in the population, there must be a sufficient rate of vertical transmission ($v>\hat v$, Corollary~\autoref{corollary:vhat}).
% TODO carry on summarizing corollaries here

The results that we derived for a deterministic well-mixed model provide a good approximation to results of simulations of a complex stochastic spatial model  (\autoref{fig:spatial}).
In this spatial model, individuals can only interact with and transmit to their neighbors.
This model demonstrates that social association between cooperation and transmission can arise in a structured population in which both types of interactions are local.
